---
title: "QBS 103 Final Project Part 3"
author: "Chichi Illoh"
date: "2023-08-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.



```{r}
setwd("/Users/Student/Desktop/QBS103/qbs103_finalproj/")

```

```{r}
meta_data = read.csv("QBS103_finalProject_metadata.csv")
#head(meta_data)
gene_data = read.csv("QBS103_finalProject_geneExpression.csv")
#head(gene_data)

```

Gene:LCN2  
Continuous variable: Age  
Categorical Variables: Sex and Disease Status  
2 more Continuous variables:ferritin, procalcitonin, fibrinogen
1 more categorical variables:ICU status
```{r}
# convert into vertical data and extract specific gene: LCN2
gene_expression = matrix(gene_data[5,2:127]) 
head(gene_expression)

# new data frame created to link variables from both data sets into one
# Adding 2 more continuous variables and one categorical variable
myData <- data.frame('Subject ID' = seq(1:126), "Gene.Expression" = gene_expression,'Age' = meta_data$age, 'Sex' = meta_data$sex, 'Disease.Status' = meta_data$disease_status, 'ICU.Status' = meta_data$icu_status, 'Ferritin' = meta_data$ferritin.ng.ml., "Procalcitonin" = meta_data$procalcitonin.ng.ml.., "Fibrinogen" = meta_data$fibrinogen)
head(myData)


myData <- myData[!(row.names(myData) %in% c("6","86","104","115")),] # filter out data that do not have exact age and filter out data where sex is unknown
# Source: https://sparkbyexamples.com/r-programming/drop-dataframe-rows-in-r/?expand_article=1


myData$Age<- as.numeric(myData$Age)# Convert age variable to numeric
myData$Gene.Expression<- as.numeric(myData$Gene.Expression) #convert gene expression variable to numeric
myData$Ferritin<- as.numeric(myData$Ferritin)
myData$Procalcitonin<- as.numeric(myData$Procalcitonin)
myData$Fibrinogen<- as.numeric(myData$Fibrinogen)
```

```{r}
data.frame('n' = c(table(myData$Sex)),
           'perc' = c(round(table(myData$Sex)/122*100,digits = 2)))

data.frame('n' = c(table(myData$Disease.Status)),
           'perc' = c(round(table(myData$Disease.Status)/122*100,digits = 2)))

data.frame('n' = c(table(myData$ICU.Status)),
           'perc' = c(round(table(myData$ICU.Status)/122*100,digits = 2)))
mean(myData$Ferritin, na.rm = TRUE)
sd(myData$Ferritin, na.rm = TRUE)

mean(myData$Procalcitonin, na.rm = TRUE)
sd(myData$Procalcitonin, na.rm = TRUE)

mean(myData$Fibrinogen, na.rm = TRUE)
sd(myData$Fibrinogen, na.rm = TRUE)
```


```{r}
#require(styler)
library(reprex)
library(dplyr)
library(magick)
library(knitr)
library(kableExtra)

# Define Table
table1 <- data.frame('Variable' = c('**Age** mean (sd)','**Sex** n (%)',
                                    "Female","Male", '**Disease Status** n (%)', 'no','yes','**ICU Status**', "Yes","No",'**Ferritin** mean (sd)','**Procalcitonin** mean (sd)','**Fibrinogen** mean (sd)'),
                     "Value" = c('61 (16)','','50 (41.0)','72 (59.0)','','98 (80.3)','24 (19.7)','','57 (46.7)','65 (53.3)','851 (1051.5)', '3 (9.9)', '519 (199.4)'))

# Print table using kable
t1 <- kable(x = table1, format = "html", caption = 'Table 1',
      col.names = c("Variable", "n = 122"),
      align = c('l','r'),escape = T) %>%
   add_indent(positions = c(3,4,6,7,9,10)) %>%
  kable_classic()
save_kable(t1, file = "t1.png")

```

```{r}
library(tidyverse)
dev.size()
hist_labels <- c("Gene Expression Distribution for ", "Gene")
hist_xlabels <- c("Gene Expression")
hist_title<- eval(bquote(expression(.(hist_labels[1])~italic(.("LCN2"))~. (hist_labels[2]))))
hist_xaxis<-eval(bquote(expression(~italic(.("LCN2"))~. (hist_xlabels[1]))))
scatter_labels <-c("Gene Expression of ", "by Age")
scatter_title <- eval(bquote(expression(.(scatter_labels[1])~italic(.("LCN2"))~. (scatter_labels[2]))))
 # Histogram for gene expression
    histogram<- ggplot(myData, aes(x=Gene.Expression)) +
  
    geom_histogram(binwidth = 39,color = "darkblue", fill = "lightblue")+
  
    labs(title = hist_title, x = hist_xaxis , y = "Number of Participants")+
      
      theme(plot.title = element_text(hjust = 0.5))
ggsave("histogram.png")  


    BlankTheme <- theme(# Remove borders and grid lines
        panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        # Define my axis
        axis.line = element_line(colour = "black", linewidth = rel(1)),
        # Set plot background to white
        plot.background = element_rect(fill = "white"),
        panel.background = element_blank(),
        legend.key = element_rect(fill = 'white'),
        # Move legend to the top
        legend.position = 'top')
# Scatterplot for gene expression and continuous covariate
    scatterplot<-ggplot(myData, aes(x = Age, y = Gene.Expression, color = Disease.Status))+
  
      geom_point(size = 2)+
      
      labs(title = scatter_title, x = "Age (yrs)", y = hist_xaxis, color = "Disease Status")+
  
      scale_color_manual(values = c("#0066CC", "#CCCCFF")) +
  
      theme(plot.title = element_text(hjust = 0.5))+
      
      BlankTheme+
      
      scale_color_discrete(labels=c('Has COVID-19', 'No COVID-19'))
ggsave("scatter.png")       
      
    
    



    BlankTheme <- theme(# Remove borders and grid lines
            panel.border = element_blank(), panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(), 
            # Define my axis
            axis.line = element_line(colour = "black", linewidth = rel(1)),
            # Set plot background to white
            plot.background = element_rect(fill = "white"),
            panel.background = element_blank(),
            legend.key = element_rect(fill = 'white'), 
            # Move legend to the top
            legend.position = 'top')
    # Box plot of gene expression separated by sex and disease status

    boxplot<-ggplot(myData, aes(x = Disease.Status, y = Gene.Expression, fill = Sex))+
  
      geom_boxplot()+ 
  
      
  
  
      labs(title = "Boxplot of Gene Expression grouped by Disease Status and Sex", x = "Disease Status", y = hist_xaxis, fill = "Sex")+
  
      BlankTheme+
      
      theme(plot.title = element_text(hjust = 0.5))+
      
      scale_x_discrete(labels = c("Has COVID-19", "No COVID-19"))+
      
      scale_fill_discrete(labels = c("Not in the ICU", "In the ICU"))+
      
      scale_fill_manual(values = c('#00CC66','#6699CC')) 
 ggsave("boxplot.png")    
    
      plot(histogram)
      plot(scatterplot)
      plot(boxplot)
      

```


```{r}
genes_11 <-gene_data[5:15,2:127]

# Calculate variance of each miRNA
variance <- apply(genes_11,MARGIN = 1,FUN = var)
# Order rows of miRNA so that highest variance in expression is on top
genes_d <- genes_11[order(variance,decreasing = T),]
# Log2-normalize data for plotting
log2.genes_d <- log2(genes_d)
#log2.genes_d <- log2.genes_d[!(row.names(log2.genes_d) %in% c("13")),]
# removing columns where ther is an infinite valu bevause heatmap will not run if ther is one
log2.genes_d <- subset(log2.genes_d, select = -106)
log2.genes_d <- subset(log2.genes_d, select = -116)
#log2.genes_d <- as.matrix(log2.genes_d)
#class(log2.genes_d)
#as.numeric(log2.genes_d[,1:126])
```




```{r}
library(pheatmap)
#generate a heatmap
#png(file="heatmap.jpg")

annotationData <- data.frame(row.names = colnames(log2.genes_d[1:10,]),
                             'Disease_Status' = c(factor(myData$Disease.Status[1:124],labels = c('COVID','non-COVID'))))
annotationData2 <- data.frame(row.names = colnames(log2.genes_d[1:10,]),
                             'ICU_Status' = c(factor(myData$ICU.Status[1:124],labels = c('in ICU','not in ICU'))))
annotationColors <- list(Status = c('COVID' = 'aquamarine4',
                                    'non-COVID' = 'deepskyblue4'),
                         ICU_Status = c('in ICU' = 'aquamarine4',
                                    'not in ICU' = 'deepskyblue4'))

# Generate heatmap
pheatmap(as.matrix(log2.genes_d[1:10,]),clustering_distance_cols = 'euclidean',
         clustering_distance_rows = 'euclidean', annotation_row =annotationData2, annotation_col = annotationData,annotation_colors = annotationColors, show_colnames = FALSE)
#dev.off()
#
```

```{r}
#generate a plot type that we did not previously discuss in class that describes your data in a new and unique way 

ggplot(myData, aes(sample = Gene.Expression)) +
  stat_qq() +
  stat_qq_line()+
  labs(title = "QQline of LCN2 Gene Expression")+
  theme(plot.title = element_text(hjust = 0.5))
ggsave("qqline.png") 
```