---
title: "QBS 103 Final Project"
author: "Chichi Illoh"
date: "2023-07-24"
output:
  html_document: default
  pdf_document: default
---

## Variables used to generate plots
Gene:LCN2  
Continuous variable: Age  
Categorical Variables: Sex and Disease Status  

The LCN2 gene encodes the protein Lipocalin 2 which plays a role in innate immunity by limiting bacterial growth through the sequestering of iron-containing siderophores. This protein is thought to be be involved in multiple cellular processes: the maintenance of skin homeostasis and suppression of invasiveness and metastasis.

In relation to COVID-19, researchers found that LCN2 is overexpressed in COVID-19 patients. They also found that this gene is linked to neutrophil and virus response activities,so higher expression of this gene leads to inflammatory responses and cilium movement.  

Sources:  

https://www.genecards.org/cgi-bin/carddisp.pl?gene=LCN2  

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7714049/  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
setwd("C:/Users/Student/Desktop/QBS103/")# set working directory

```

```{r}
library(tidyverse)

meta_data = read.csv("QBS103_finalProject_metadata.csv")
#head(meta_data)
gene_data = read.csv("QBS103_finalProject_geneExpression.csv")
#head(gene_data)

# convert into vertical data and extract specific gene: LCN2
gene_expression = matrix(gene_data[5,2:127]) 
head(gene_expression)

# new data frame created to link variables from both data sets into one
myData <- data.frame('Subject ID' = seq(1:126), "Gene.Expression" = gene_expression,'Age' = meta_data$age, 'Sex' = meta_data$sex, 'Disease.Status' = meta_data$disease_status, 'ICU.Status' = meta_data$icu_status)
head(myData)


myData <- myData[!(row.names(myData) %in% c("6","86","104","115")),] # filter out data that do not have exact age and filter out data where sex is unknown
# Source: https://sparkbyexamples.com/r-programming/drop-dataframe-rows-in-r/?expand_article=1


myData$Age<- as.numeric(myData$Age)# Convert age variable to numeric
myData$Gene.Expression<- as.numeric(myData$Gene.Expression) #convert gene expression variable to numeric
```

```{r}

# Histogram for gene expression
ggplot(myData, aes(x=Gene.Expression)) +
  
  geom_histogram(binwidth = 39,color = "darkblue", fill = "lightblue")+
  
  labs(title = "LCN2 Gene Expression Distribution", x = "LCN2 Gene Expression", y = "Count")
  
  
```

```{r}
BlankTheme <- theme(# Remove borders and grid lines
        panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        # Define my axis
        axis.line = element_line(colour = "black", linewidth = rel(1)),
        # Set plot background to white
        plot.background = element_rect(fill = "white"),
        panel.background = element_blank(),
        legend.key = element_rect(fill = 'white'),
        # Move legend to the top
        legend.position = 'top')
# Scatterplot for gene expression and continuous covariate
ggplot(myData, aes(x = Age, y = Gene.Expression, color = Disease.Status))+
  
  geom_point(size = 2)+
  
  labs(title = "Gene Expression of LCN2 by Age", x = "Age (yrs)", y = " LCN2 Gene Expression (AU)", color = "Disease Status")+
  
  scale_color_manual(values = c("#0066CC", "#CCCCFF")) +
  
  BlankTheme

```

```{r}
BlankTheme <- theme(# Remove borders and grid lines
        panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        # Define my axis
        axis.line = element_line(colour = "black", linewidth = rel(1)),
        # Set plot background to white
        plot.background = element_rect(fill = "white"),
        panel.background = element_blank(),
        legend.key = element_rect(fill = 'white'), 
        # Move legend to the top
        legend.position = 'top')
# Box plot of gene expression separated by sex and disease status

ggplot(myData, aes(x = Disease.Status, y = Gene.Expression, color = Sex))+
  
  geom_boxplot()+
  
  scale_color_manual(values = c('#00CC66','#6699CC')) +
  
  #scales_x_discrete
  
  labs(title = "Gene Expression of LCN2 grouped by Disease status\n and categorized by Sex", x = "Disease Status", y = "LCN2 Gene Expression (AU)")+
  
  BlankTheme

```

```{r}
meta_data = read.csv("QBS103_finalProject_metadata.csv")
#head(meta_data)
gene_data = read.csv("QBS103_finalProject_geneExpression.csv")
#head(gene_data)



```


```{r}
stats_plts <- function(data_frame, gene_names,contin_covariate, categorical_var1,categorical_var2)
  {
  # reading the data
  library(tidyverse)
  for (gene in gene_names){
  # convert into vertical data and extract specific gene from gene_names list
    i = which(data_frame$X == gene)
    gene_expression = matrix(data_frame[i,2:127]) 
    head(gene_expression)

    # new data frame created to link variables from both data sets into one
    newData <- data.frame('Subject ID' = seq(1:126), "Gene.Expression" = gene_expression,'Continuous_Var' = contin_covariate, 'Categorical_Var1' = categorical_var1, 'Categorical_Var2' = categorical_var2)
    head(newData)

#----------MIGHT HAVE TO REMOVE BELOW-----------------------
    newData <- newData[!(row.names(newData) %in% c("6","86","104","115")),] # filter out data that do     not have exact age and filter out data where sex is unknown
    # Source: https://sparkbyexamples.com/r-programming/drop-dataframe-rows-in-r/?expand_article=1


    newData$Continuous_Var<- as.numeric(newData$Continuous_Var)# Convert age variable to numeric
    newData$Gene.Expression<- as.numeric(newData$Gene.Expression) #convert gene expression variable      to numeric
  
  
  
  
  
  
  # Histogram for gene expression
    histogram<- ggplot(newData, aes(x=Gene.Expression)) +
  
    geom_histogram(binwidth = 39,color = "darkblue", fill = "lightblue")+
  
    labs(title = "LCN2 Gene Expression Distribution", x = "LCN2 Gene Expression", y = "Count")
  


    BlankTheme <- theme(# Remove borders and grid lines
        panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        # Define my axis
        axis.line = element_line(colour = "black", linewidth = rel(1)),
        # Set plot background to white
        plot.background = element_rect(fill = "white"),
        panel.background = element_blank(),
        legend.key = element_rect(fill = 'white'),
        # Move legend to the top
        legend.position = 'top')
# Scatterplot for gene expression and continuous covariate
    scatterplot<-ggplot(newData, aes(x = Continuous_Var, y = Gene.Expression, color = Categorical_Var1))+
  
      geom_point(size = 2)+
  
      labs(title = "Gene Expression of e", x = "cont var", y = " Gene Expression (AU)", color = "Categorical_Var1")+
  
      scale_color_manual(values = c("#0066CC", "#CCCCFF")) +
  
      BlankTheme



    BlankTheme <- theme(# Remove borders and grid lines
            panel.border = element_blank(), panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(), 
            # Define my axis
            axis.line = element_line(colour = "black", linewidth = rel(1)),
            # Set plot background to white
            plot.background = element_rect(fill = "white"),
            panel.background = element_blank(),
            legend.key = element_rect(fill = 'white'), 
            # Move legend to the top
            legend.position = 'top')
    # Box plot of gene expression separated by sex and disease status

    boxplot<-ggplot(newData, aes(x = Categorical_Var1, y = Gene.Expression, color = Categorical_Var2))+
  
      geom_boxplot()+
  
      scale_color_manual(values = c('#00CC66','#6699CC')) +
  
      #scales_x_discrete
  
      labs(title = "Gene Expression of gene grouped by Categorical_Var1 \n and categorized by Categorical_Var2", x = "Categorical_Var1", y = " Gene Expression (AU)")+
  
      BlankTheme
      plot(histogram)
      plot(scatterplot)
      plot(boxplot)
}
}
```

```{r}

stats_plts(gene_data, gene_names = list("LCN2","CD24","PRTN3"), meta_data$age, meta_data$disease_status,meta_data$icu_status)

```

